---
title: "Modeling"
author: "Leonor Brito"
date: "2024-09-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# packages
```{r}

library(lavaan)
library(blavaan)
library(ggplot2)
library(purrr)
library(OpenMx)
library(dplyr)

```

```{r}

load("/Users/leonorbrito/Documents/Master/Thesis/Code/clean_df.RData")

```

Based on exploratory data analysis, a shorter list of variables was selected for data analysis and modeling.

```{r}

final_df <- clean_df[, c("year", "area", "cereal_yield", "irrgation_efficiency", "pesticides", "organic_ag", "enteric_fermentation", "manure_ch4", "farm_energy_n20", "gdp_capita", "employment_ratio", "temp_change", "percipitation", "floods", "nue", "kue", "pue")]

save(final_df, file = "final_df.RData")

```

# Assumptions / Analysis

## Correlation

```{r}

vars <- final_df[, 4:ncol(final_df)]

corr_results <- rcorr(as.matrix(vars))

# access the matrix of correlation coefficients
cor_matrix <- corr_results$r

custom_labels <- c(
  "irrgation_efficiency" = "Irrigation Efficiency",
  "pesticides" = "Pesticide Use",
  "organic_ag" = "Organic Agriculture",
  "enteric_fermentation" = "Enteric Fermentation",
  "manure_ch4" = "Manure Management",
  "farm_energy_n20" = "On-Farm Energy Use",
  "gdp_capita" = "GDP per Capita",
  "employment_ratio" = "Employment Ratio",
  "temp_change" = "Temperature Change",
  "percipitation" = "Precipitation",
  "floods" = "Floods",
  "nue" = "Nitrogen Use Efficiency",
  "kue" = "Potassium Use Efficiency",
  "pue" = "Phosphorus Use Efficiency"
)

# apply custom labels to correlation matrix
colnames(cor_matrix) <- custom_labels[colnames(cor_matrix)]
rownames(cor_matrix) <- custom_labels[rownames(cor_matrix)]

# access the matrix of p-values
p_matrix <- corr_results$P

# visualize the correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, tl.cex = 0.6, number.cex = 0.5, addCoef.col = "black")

# save the correlation plot as a PNG file
png("correlation.png", width = 1400, height = 600, res = 300)

# generate the correlation plot
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, tl.cex = 0.6, addCoef.col = NULL)

# close
dev.off()

```

## EFA

```{r}

# perform EFA with an initial guess of the number of factors
initial_efa <- fa(vars, nfactors = 6, rotate = "varimax") 

# get factor loadings and convert to a tidy format
loadings_df <- as.data.frame(initial_efa$loadings[])
loadings_df <- loadings_df %>%
  rownames_to_column("Variable") %>%
  pivot_longer(-Variable, names_to = "Factor", values_to = "Loading")

# filter for the first two factors
loadings_df_filtered <- loadings_df %>% filter(Factor %in% c("MR1", "MR2"))

# customize variable names for x-axis labels
loadings_df_filtered$Variable <- factor(loadings_df_filtered$Variable, levels = unique(loadings_df_filtered$Variable))

# set the variable names as factors with the order from custom_labels
loadings_df_filtered$Variable <- factor(loadings_df_filtered$Variable, levels = names(custom_labels), labels = custom_labels)

# separate data for each factor and reorder variables by loading
loadings_df_ordered <- loadings_df_filtered %>%
  group_by(Factor) %>%
  arrange(Factor, desc(Loading)) %>%
  mutate(Variable = factor(Variable, levels = Variable)) %>%
  ungroup()

# add a new column for color
loadings_df_plot <- loadings_df_ordered %>%
  group_by(Factor) %>%
  mutate(Color = ifelse(row_number() <= 3, "Top 3", "Others")) %>%
  ungroup()

# plot the factor loadings for the first two factors as horizontal bar graphs
ggplot(loadings_df_plot, aes(x = Loading, y = Variable, fill = Color)) +
  geom_bar(stat = "identity", position = position_dodge(), show.legend = FALSE) +
  facet_wrap(~ Factor) +
  labs(x = "Loading", y = NULL) + 
  scale_fill_manual(values = c("Top 3" = "#1f78b4", "Others" = "gray")) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10, hjust = 1), 
    axis.text.x = element_text(size = 10), 
    axis.title.x = element_text(size = 12), 
    panel.grid.major.x = element_line(color = "grey80"), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none", 
    strip.background = element_blank(), 
    strip.text = element_text(size = 12, face = "bold") 
  )

ggsave("EFA.png", width = 14, height = 6, dpi = 300, bg = "white")

```

## Variable Transformations

```{r}

plot_transformations <- function(data_vector) {
  # Create a data frame from the vector for easier manipulation
  data_df <- data.frame(Value = data_vector)
  
  # Remove NA values for transformation and plotting
  data_df <- data_df %>% filter(!is.na(Value))
  
  # Define transformations
  transformations <- list(
    "Original" = data_df$Value,
    "Log" = log(data_df$Value),
    "LogLog" = log(log(data_df$Value[data_df$Value > 1])),
    "Square Root" = sqrt(data_df$Value),
    "Cube Root" = data_df$Value^(1/3)
  )
  
  # Plotting loop
  par(mfrow = c(5, 2), mar = c(3, 3, 2, 1))  # Adjust margins and layout
  for (trans_name in names(transformations)) {
    # Histogram
    hist_data <- transformations[[trans_name]]
    hist_data <- hist_data[!is.infinite(hist_data) & !is.na(hist_data)]  # Remove Inf and NA values
    
    hist(hist_data, main = paste(trans_name, "Distribution"), xlab = trans_name, col = "blue", breaks = 30)
    
    # Boxplot
    boxplot(hist_data, horizontal = TRUE, main = paste(trans_name, "Boxplot"), xlab = trans_name)
  }
}

plot_transformations(final_df$pesticides)

```

Apply transformations
```{r}

# Log transformation
final_df$log_manure <- ifelse(is.na(final_df$manure_ch4), NA, log(final_df$manure_ch4))
final_df$log_pesticides <- ifelse(is.na(final_df$pesticides), NA, log(final_df$pesticides))
final_df$log_farm_energy_n20 <- ifelse(is.na(final_df$farm_energy_n20), NA, log(final_df$farm_energy_n20))

# Square root transformation
final_df$sqrt_enteric <- ifelse(is.na(final_df$enteric_fermentation), NA, sqrt(final_df$enteric_fermentation))
final_df$sqrt_employment <- ifelse(is.na(final_df$employment_ratio), NA, sqrt(final_df$employment_ratio))
final_df$sqrt_gdp_capita <- ifelse(is.na(final_df$gdp_capita), NA, sqrt(final_df$gdp_capita))
final_df$sqrt_irrigation <- ifelse(is.na(final_df$irrgation_efficiency), NA, sqrt(final_df$irrgation_efficiency))
final_df$sqrt_organic <- ifelse(is.na(final_df$organic_ag), NA, sqrt(final_df$organic_ag))


```

Save df with variable transformations
```{r}

transformed_df <- final_df
save(transformed_df, file = "transformed_df.RData")

```


# Scaling
```{r}

data_sem <- final_df[, !(names(final_df) %in% c("Country.Group.Code", "Country.Group", "gdp_category"))]
country_data <- data_sem[,c("area"), drop = FALSE]
data_to_scale <- data_sem[, !(names(data_sem) %in% c("area"))]
data_scaled <- as.data.frame(lapply(data_to_scale, scale))

data_for_sem <- cbind(data_scaled, country_data)

```

# CFA

```{r}

cfa <- '

  NutrientUseEfficiency =~ nue + pue + kue
  AgriculturalEmissions =~ log_manure + sqrt_enteric + log_farm_energy_n20
  
'

# fit cfa model and output summary
fit_cfa <- sem(cfa, data = data_for_sem, cluster = 'area', missing = "FIML", estimator = "MLR")
summary(fit_cfa, fit.measures=TRUE)

```


# Baseline SEM Structure
```{r}

baseline_model <- '

  # Measurement Model
  NutrientUseEfficiency =~ 1*nue + pue + kue
  AgriculturalEmissions =~ 1*log_manure + sqrt_enteric
  
  # Structural Model
  # Direct effects
  cereal_yield ~ b1*NutrientUseEfficiency + b2*log_pesticides + b3*percipitation + b4*AgriculturalEmissions + b5*temp_change + b6*sqrt_gdp_capita + b7*sqrt_irrigation + b8*employment_ratio + b9*sqrt_organic
  
  # Mediator path
  NutrientUseEfficiency ~ m1*log_pesticides + m2*floods + m3*percipitation + m4*sqrt_organic 
  
  # Indirect effects through mediated path
  indirect_log_pesticides := m1 * b1
  indirect_floods := m2 * b1
  indirect_percipitation := m3 * b1
  indirect_sqrt_organic := m4 * b1
  

  # Total effects through mediated path
  total_effect_log_pesticides := b2 + indirect_log_pesticides
  total_effect_floods := indirect_floods
  total_effect_percipitation := b3 + indirect_percipitation
  total_effect_sqrt_organic := b9 + indirect_sqrt_organic

'

# fit model and output summary
fit_baseline_model <- sem(baseline_model, data = data_for_sem, cluster = 'area', missing = "FIML", estimator = "MLR")
summary(fit_baseline_model, fit.measures=TRUE)

```

# Modeling with time

## Adding year as variable
```{r}

model_with_year <- '

  # Measurement Model
  NutrientUseEfficiency =~ 1*nue + pue + kue
  AgriculturalEmissions =~ 1*log_manure + sqrt_enteric + log_farm_energy_n20
  
  # Structural Model
  # Direct effects
  cereal_yield ~ b1*NutrientUseEfficiency + b2*log_pesticides + b3*percipitation + b4*AgriculturalEmissions + b5*temp_change + b6*sqrt_gdp_capita + b7*sqrt_irrigation + b8*employment_ratio + b9*sqrt_organic + b10*year
  
  # Mediator path
  NutrientUseEfficiency ~ m1*log_pesticides + m2*floods + m3*percipitation + m4*sqrt_organic
  
  # Indirect effects through mediated path
  indirect_log_pesticides := m1 * b1
  indirect_floods := m2 * b1
  indirect_percipitation := m3 * b1
  indirect_sqrt_organic := m4 * b1

  # Total effects through mediated path
  total_effect_log_pesticides := b2 + indirect_log_pesticides
  total_effect_floods := indirect_floods
  total_effect_percipitation := b3 + indirect_percipitation
  total_effect_sqrt_organic := b9 + indirect_sqrt_organic

'

# fit model and output summary
fit_model_year <- sem(model_with_year, data = data_for_sem, cluster = 'area', missing = "fiml", estimator = "MLR")
summary(fit_model_year, fit.measures=TRUE)

```


## Residuals from time series

Arima time series
```{r}

# fit ARIMA time series on cereal yield
arima_model <- auto.arima(final_df$cereal_yield, xreg = final_df$year)

# calculate and store residuals
data_for_sem$cereal_yield_arima <- residuals(arima_model)

# scale residuals
data_for_sem$cereal_yield_arima <- scale(data_for_sem$cereal_yield_arima)

```

```{r}

# plot residuals for the time series model
ggplot(data_for_sem, aes(x = year, y = cereal_yield_arima)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "",
       x = "Year",
       y = "Residuals") +
  theme_minimal()

ggsave("arima.png", width = 14, height = 6, dpi = 300, bg = "white")

```


re-fit with residuals from time series
```{r}

model_arima <- '

  # Measurement Model
  NutrientUseEfficiency =~ 1*nue + pue + kue
  AgriculturalEmissions =~ 1*log_manure + sqrt_enteric + log_farm_energy_n20
  
  # Structural Model
  # Direct effects
  cereal_yield_arima ~ b1*NutrientUseEfficiency + b2*log_pesticides + b3*percipitation + b4*AgriculturalEmissions + b5*temp_change + b6*sqrt_gdp_capita + b7*sqrt_irrigation + b8*employment_ratio + b9*sqrt_organic
  
  # Mediator path
  NutrientUseEfficiency ~ m1*log_pesticides + m2*floods + m3*percipitation + m4*sqrt_organic 
  
  # Indirect effects through mediated path
  indirect_log_pesticides := m1 * b1
  indirect_floods := m2 * b1
  indirect_percipitation := m3 * b1
  indirect_sqrt_organic := m4 * b1
  

  # Total effects through mediated path
  total_effect_log_pesticides := b2 + indirect_log_pesticides
  total_effect_floods := indirect_floods
  total_effect_percipitation := b3 + indirect_percipitation
  total_effect_sqrt_organic := b9 + indirect_sqrt_organic

'
# fit model and output summary
fit_model_arima <- sem(model_arima, data = data_for_sem, cluster = 'area', missing = "fiml", estimator='MLR')
summary(fit_model_arima, fit.measures=TRUE)

```

## Residuals from linear regression

regression on year
```{r}

# regress yield on year
model_year <- lm(cereal_yield ~ year, data = final_df)

# calculate and store residuals
data_for_sem$cereal_yield_res <- resid(model_year)

#scale residuals
data_for_sem$cereal_yield_res <- scale(data_for_sem$cereal_yield_res)

```

```{r}

# plot residuals for the time series model
ggplot(data_for_sem, aes(x = year, y = cereal_yield_res)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "",
       x = "Year",
       y = "Residuals") +
  theme_minimal()

ggsave("linear_reg.png", width = 14, height = 6, dpi = 300, bg = "white")

```


re-fit with residuals from regression on year
```{r}

model_linear_reg <- '

  # Measurement Model
  NutrientUseEfficiency =~ 1*nue + pue + kue
  AgriculturalEmissions =~ 1*log_manure + sqrt_enteric + log_farm_energy_n20
  
  # Structural Model
  # Direct effects
  cereal_yield_res ~ b1*NutrientUseEfficiency + b2*log_pesticides + b3*percipitation + b4*AgriculturalEmissions + b5*temp_change + b6*sqrt_gdp_capita + b7*sqrt_irrigation + b8*employment_ratio + b9*sqrt_organic
  
  # Mediator path
  NutrientUseEfficiency ~ m1*log_pesticides + m2*floods + m3*percipitation + m4*sqrt_organic 
  
  # Indirect effects through mediated path
  indirect_log_pesticides := m1 * b1
  indirect_floods := m2 * b1
  indirect_percipitation := m3 * b1
  indirect_sqrt_organic := m4 * b1
  

  # Total effects through mediated path
  total_effect_log_pesticides := b2 + indirect_log_pesticides
  total_effect_floods := indirect_floods
  total_effect_percipitation := b3 + indirect_percipitation
  total_effect_sqrt_organic := b9 + indirect_sqrt_organic

'

# fit model and output summary
fit_model_reg <- sem(model_linear_reg, data = data_for_sem, cluster = 'area', missing = "fiml", estimator='MLR')
summary(fit_model_reg, fit.measures=TRUE)

```

Likelihood Ratio Test (LRT) to compare multiple SEM models
```{r}

comparison <- lavTestLRT(fit_baseline_model, fit_model_arima, fit_model_reg, fit_model_year)
print(comparison)

```
