---
title: "Modeling"
author: "Leonor Brito"
date: "2024-09-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# packages
```{r}

library(lavaan)
library(blavaan)
library(ggplot2)
library(purrr)
library(OpenMx)
library(dplyr)

```

```{r}

load("/Users/leonorbrito/Documents/Master/Thesis/Code/clean_df.RData")

```

Based on exploratory data analysis, a shorter list of variables was selected for data analysis and modeling.

```{r}

final_df <- clean_df[, c("year", "area", "cereal_yield", "irrgation_efficiency", "pesticides", "organic_ag", "enteric_fermentation", "manure_ch4", "farm_energy_n20", "gdp_capita", "employment_ratio", "temp_change", "percipitation", "floods", "nue", "kue", "pue")]

save(final_df, file = "final_df.RData")

```

# Assumptions / Analysis

## Correlation

```{r}

vars <- final_df[, 4:ncol(final_df)]

corr_results <- rcorr(as.matrix(vars))

# access the matrix of correlation coefficients
cor_matrix <- corr_results$r

custom_labels <- c(
  "irrgation_efficiency" = "Irrigation Efficiency",
  "pesticides" = "Pesticide Use",
  "organic_ag" = "Organic Agriculture",
  "enteric_fermentation" = "Enteric Fermentation",
  "manure_ch4" = "Manure Management",
  "farm_energy_n20" = "On-Farm Energy Use",
  "gdp_capita" = "GDP per Capita",
  "employment_ratio" = "Employment Ratio",
  "temp_change" = "Temperature Change",
  "percipitation" = "Precipitation",
  "floods" = "Floods",
  "nue" = "Nitrogen Use Efficiency",
  "kue" = "Potassium Use Efficiency",
  "pue" = "Phosphorus Use Efficiency"
)

# apply custom labels to correlation matrix
colnames(cor_matrix) <- custom_labels[colnames(cor_matrix)]
rownames(cor_matrix) <- custom_labels[rownames(cor_matrix)]

# access the matrix of p-values
p_matrix <- corr_results$P

# visualize the correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, tl.cex = 0.6, number.cex = 0.5, addCoef.col = "black")

# save the correlation plot as a PNG file
png("correlation.png", width = 1400, height = 600, res = 300)

# generate the correlation plot
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, tl.cex = 0.6, addCoef.col = NULL)

# close
dev.off()

```

## EFA

```{r}

# perform EFA with an initial guess of the number of factors
initial_efa <- fa(vars, nfactors = 6, rotate = "varimax") 

# get factor loadings and convert to a tidy format
loadings_df <- as.data.frame(initial_efa$loadings[])
loadings_df <- loadings_df %>%
  rownames_to_column("Variable") %>%
  pivot_longer(-Variable, names_to = "Factor", values_to = "Loading")

# filter for the first two factors
loadings_df_filtered <- loadings_df %>% filter(Factor %in% c("MR1", "MR2"))

# customize variable names for x-axis labels
loadings_df_filtered$Variable <- factor(loadings_df_filtered$Variable, levels = unique(loadings_df_filtered$Variable))

# set the variable names as factors with the order from custom_labels
loadings_df_filtered$Variable <- factor(loadings_df_filtered$Variable, levels = names(custom_labels), labels = custom_labels)

# separate data for each factor and reorder variables by loading
loadings_df_ordered <- loadings_df_filtered %>%
  group_by(Factor) %>%
  arrange(Factor, desc(Loading)) %>%
  mutate(Variable = factor(Variable, levels = Variable)) %>%
  ungroup()

# add a new column for color
loadings_df_plot <- loadings_df_ordered %>%
  group_by(Factor) %>%
  mutate(Color = ifelse(row_number() <= 3, "Top 3", "Others")) %>%
  ungroup()

# plot the factor loadings for the first two factors as horizontal bar graphs
ggplot(loadings_df_plot, aes(x = Loading, y = Variable, fill = Color)) +
  geom_bar(stat = "identity", position = position_dodge(), show.legend = FALSE) +
  facet_wrap(~ Factor) +
  labs(x = "Loading", y = NULL) + 
  scale_fill_manual(values = c("Top 3" = "#1f78b4", "Others" = "gray")) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10, hjust = 1), 
    axis.text.x = element_text(size = 10), 
    axis.title.x = element_text(size = 12), 
    panel.grid.major.x = element_line(color = "grey80"), 
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none", 
    strip.background = element_blank(), 
    strip.text = element_text(size = 12, face = "bold") 
  )

ggsave("EFA.png", width = 14, height = 6, dpi = 300, bg = "white")

```

## Variable Transformations

```{r}

plot_transformations <- function(data_vector) {
  # Create a data frame from the vector for easier manipulation
  data_df <- data.frame(Value = data_vector)
  
  # Remove NA values for transformation and plotting
  data_df <- data_df %>% filter(!is.na(Value))
  
  # Define transformations
  transformations <- list(
    "Original" = data_df$Value,
    "Log" = log(data_df$Value),
    "LogLog" = log(log(data_df$Value[data_df$Value > 1])),
    "Square Root" = sqrt(data_df$Value),
    "Cube Root" = data_df$Value^(1/3)
  )
  
  # Plotting loop
  par(mfrow = c(5, 2), mar = c(3, 3, 2, 1))  # Adjust margins and layout
  for (trans_name in names(transformations)) {
    # Histogram
    hist_data <- transformations[[trans_name]]
    hist_data <- hist_data[!is.infinite(hist_data) & !is.na(hist_data)]  # Remove Inf and NA values
    
    hist(hist_data, main = paste(trans_name, "Distribution"), xlab = trans_name, col = "blue", breaks = 30)
    
    # Boxplot
    boxplot(hist_data, horizontal = TRUE, main = paste(trans_name, "Boxplot"), xlab = trans_name)
  }
}

plot_transformations(final_df$pesticides)

```

Apply transformations
```{r}

# Log transformation
final_df$log_manure <- ifelse(is.na(final_df$manure_ch4), NA, log(final_df$manure_ch4))
final_df$log_pesticides <- ifelse(is.na(final_df$pesticides), NA, log(final_df$pesticides))
final_df$log_farm_energy_n20 <- ifelse(is.na(final_df$farm_energy_n20), NA, log(final_df$farm_energy_n20))

# Square root transformation
final_df$sqrt_enteric <- ifelse(is.na(final_df$enteric_fermentation), NA, sqrt(final_df$enteric_fermentation))
final_df$sqrt_employment <- ifelse(is.na(final_df$employment_ratio), NA, sqrt(final_df$employment_ratio))
final_df$sqrt_gdp_capita <- ifelse(is.na(final_df$gdp_capita), NA, sqrt(final_df$gdp_capita))
final_df$sqrt_irrigation <- ifelse(is.na(final_df$irrgation_efficiency), NA, sqrt(final_df$irrgation_efficiency))
final_df$sqrt_organic <- ifelse(is.na(final_df$organic_ag), NA, sqrt(final_df$organic_ag))


```

Save df with variable transformations
```{r}

transformed_df <- final_df
save(transformed_df, file = "transformed_df.RData")

```


# Scaling
```{r}

data_sem <- final_df[, !(names(final_df) %in% c("Country.Group.Code", "Country.Group", "gdp_category"))]
country_data <- data_sem[,c("area"), drop = FALSE]
data_to_scale <- data_sem[, !(names(data_sem) %in% c("area"))]
data_scaled <- as.data.frame(lapply(data_to_scale, scale))

data_for_sem <- cbind(data_scaled, country_data)

```

# CFA

```{r}

cfa <- '

  NutrientUseEfficiency =~ nue + pue + kue
  AgriculturalEmissions =~ log_manure + sqrt_enteric + log_farm_energy_n20
  
'

# fit cfa model and output summary
fit_cfa <- sem(cfa, data = data_for_sem, cluster = 'area', missing = "FIML", estimator = "MLR")
summary(fit_cfa, fit.measures=TRUE)

```


# Baseline SEM Structure
```{r}

baseline_model <- '

  # Measurement Model
  NutrientUseEfficiency =~ 1*nue + pue + kue
  AgriculturalEmissions =~ 1*log_manure + sqrt_enteric + log_farm_energy_n20
  
  # Structural Model
  # Direct effects
  cereal_yield ~ b1*NutrientUseEfficiency + b2*log_pesticides + b3*percipitation + b4*AgriculturalEmissions + b5*temp_change + b6*sqrt_gdp_capita + b7*sqrt_irrigation + b8*employment_ratio + b9*sqrt_organic
  
  # Mediator path
  NutrientUseEfficiency ~ m1*log_pesticides + m2*floods + m3*percipitation + m4*sqrt_organic 
  
  # Indirect effects through mediated path
  indirect_log_pesticides := m1 * b1
  indirect_floods := m2 * b1
  indirect_percipitation := m3 * b1
  indirect_sqrt_organic := m4 * b1
  

  # Total effects through mediated path
  total_effect_log_pesticides := b2 + indirect_log_pesticides
  total_effect_floods := indirect_floods
  total_effect_percipitation := b3 + indirect_percipitation
  total_effect_sqrt_organic := b9 + indirect_sqrt_organic

'

# fit model and output summary
fit_baseline_model <- sem(baseline_model, data = data_for_sem, cluster = 'area', missing = "FIML", estimator = "MLR")
summary(fit_baseline_model, fit.measures=TRUE)

```

# Modeling with time

## Adding year as variable
```{r}

model_with_year <- '

  # Measurement Model
  NutrientUseEfficiency =~ 1*nue + pue + kue
  AgriculturalEmissions =~ 1*log_manure + sqrt_enteric + log_farm_energy_n20
  
  # Structural Model
  # Direct effects
  cereal_yield ~ b1*NutrientUseEfficiency + b2*log_pesticides + b3*percipitation + b4*AgriculturalEmissions + b5*temp_change + b6*sqrt_gdp_capita + b7*sqrt_irrigation + b8*employment_ratio + b9*sqrt_organic + b10*year
  
  # Mediator path
  NutrientUseEfficiency ~ m1*log_pesticides + m2*floods + m3*percipitation + m4*sqrt_organic
  
  # Indirect effects through mediated path
  indirect_log_pesticides := m1 * b1
  indirect_floods := m2 * b1
  indirect_percipitation := m3 * b1
  indirect_sqrt_organic := m4 * b1

  # Total effects through mediated path
  total_effect_log_pesticides := b2 + indirect_log_pesticides
  total_effect_floods := indirect_floods
  total_effect_percipitation := b3 + indirect_percipitation
  total_effect_sqrt_organic := b9 + indirect_sqrt_organic

'

# fit model and output summary
fit_model_year <- sem(model_with_year, data = data_for_sem, cluster = 'area', missing = "fiml", estimator = "MLR")
summary(fit_model_year, fit.measures=TRUE)

```


## Residuals from time series

Arima time series
```{r}

# fit ARIMA time series on cereal yield
arima_model <- auto.arima(final_df$cereal_yield, xreg = final_df$year)

# calculate and store residuals
data_for_sem$cereal_yield_arima <- residuals(arima_model)

# scale residuals
data_for_sem$cereal_yield_arima <- scale(data_for_sem$cereal_yield_arima)

```

```{r}

# plot residuals for the time series model
ggplot(data_for_sem, aes(x = year, y = cereal_yield_arima)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "",
       x = "Year",
       y = "Residuals") +
  theme_minimal()

ggsave("arima.png", width = 14, height = 6, dpi = 300, bg = "white")

```


re-fit with residuals from time series
```{r}

model_arima <- '

  # Measurement Model
  NutrientUseEfficiency =~ 1*nue + pue + kue
  AgriculturalEmissions =~ 1*log_manure + sqrt_enteric + log_farm_energy_n20
  
  # Structural Model
  # Direct effects
  cereal_yield_arima ~ b1*NutrientUseEfficiency + b2*log_pesticides + b3*percipitation + b4*AgriculturalEmissions + b5*temp_change + b6*sqrt_gdp_capita + b7*sqrt_irrigation + b8*employment_ratio + b9*sqrt_organic
  
  # Mediator path
  NutrientUseEfficiency ~ m1*log_pesticides + m2*floods + m3*percipitation + m4*sqrt_organic 
  
  # Indirect effects through mediated path
  indirect_log_pesticides := m1 * b1
  indirect_floods := m2 * b1
  indirect_percipitation := m3 * b1
  indirect_sqrt_organic := m4 * b1
  

  # Total effects through mediated path
  total_effect_log_pesticides := b2 + indirect_log_pesticides
  total_effect_floods := indirect_floods
  total_effect_percipitation := b3 + indirect_percipitation
  total_effect_sqrt_organic := b9 + indirect_sqrt_organic

'
# fit model and output summary
fit_model_arima <- sem(model_arima, data = data_for_sem, cluster = 'area', missing = "fiml", estimator='MLR')
summary(fit_model_arima, fit.measures=TRUE)

```

## Residuals from linear regression

regression on year
```{r}

# regress yield on year
model_year <- lm(cereal_yield ~ year, data = final_df)

# calculate and store residuals
data_for_sem$cereal_yield_res <- resid(model_year)

#scale residuals
data_for_sem$cereal_yield_res <- scale(data_for_sem$cereal_yield_res)

```

```{r}

# plot residuals for the time series model
ggplot(data_for_sem, aes(x = year, y = cereal_yield_res)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "",
       x = "Year",
       y = "Residuals") +
  theme_minimal()

ggsave("linear_reg.png", width = 14, height = 6, dpi = 300, bg = "white")

```


re-fit with residuals from regression on year
```{r}

model_linear_reg <- '

  # Measurement Model
  NutrientUseEfficiency =~ 1*nue + pue + kue
  AgriculturalEmissions =~ 1*log_manure + sqrt_enteric + log_farm_energy_n20
  
  # Structural Model
  # Direct effects
  cereal_yield_res ~ b1*NutrientUseEfficiency + b2*log_pesticides + b3*percipitation + b4*AgriculturalEmissions + b5*temp_change + b6*sqrt_gdp_capita + b7*sqrt_irrigation + b8*employment_ratio + b9*sqrt_organic
  
  # Mediator path
  NutrientUseEfficiency ~ m1*log_pesticides + m2*floods + m3*percipitation + m4*sqrt_organic 
  
  # Indirect effects through mediated path
  indirect_log_pesticides := m1 * b1
  indirect_floods := m2 * b1
  indirect_percipitation := m3 * b1
  indirect_sqrt_organic := m4 * b1
  

  # Total effects through mediated path
  total_effect_log_pesticides := b2 + indirect_log_pesticides
  total_effect_floods := indirect_floods
  total_effect_percipitation := b3 + indirect_percipitation
  total_effect_sqrt_organic := b9 + indirect_sqrt_organic

'

# fit model and output summary
fit_model_reg <- sem(model_linear_reg, data = data_for_sem, cluster = 'area', missing = "fiml", estimator='MLR')
summary(fit_model_reg, fit.measures=TRUE)

```

Likelihood Ratio Test (LRT) to compare multiple SEM models
```{r}

comparison <- lavTestLRT(fit_baseline_model, fit_model_arima, fit_model_reg, fit_model_year)
print(comparison)

```


---

The following code was NOT USED OR PURSUED in this study. It represents a potential approach to multi-level SEM modeling that was unsuccessful. I have left some of the work here below as I refer to it in the Discussion as part of future work.

The idea was to split the data into 2 (or more) time periods, and build two SEM models: 
1. Free model: in this model, two SEM structures are built for each time period. For each time period, the variables have different coefficients. This assumes that each variable has a difference effect on yield before 2006 and after 2006 (i.e. time is important)
2. Constrained model: in this model, two SEM structures are built for each time period, but for each time period, the variables have the same coefficients. This assumes that time does not influence the effects of each variable on yield.

The two models can be compared using Likelihood Ratio test to see which one performs best with the data.

Split data into 2 time periods 
```{r}

final_df <- final_df %>%
  mutate(time_period = ifelse(year %in% 2006:2013, "t1", "t2"))

# relevant variables excluding floods and percipitation
relevant_vars <- c("nue", "pue", "kue", "manure_direct", "farm_energy_n20", "enteric_fermentation",
                   "pesticides", "gdp_capita", "temp_change", "employment_ratio", "water_withdrawal", "cereal_yield")

# aggregate data by country, time period, and the relevant variables
DF <- final_df %>%
  select(area, time_period, all_of(relevant_vars), floods, percipitation) %>%
  group_by(area, time_period) %>%
  summarise(across(.cols = all_of(relevant_vars), .fns = list(mean = mean, sd = sd), na.rm = TRUE),
            floods = first(floods), percipitation = first(percipitation),
            .groups = 'drop')

# check for duplicates
duplicates <- DF %>%
  group_by(area, time_period) %>%
  filter(n() > 1)

if(nrow(duplicates) > 0) {
  print("Duplicates found:")
  print(duplicates)
} else {
  print("No duplicates found.")
}

DF_wide <- DF %>%
  pivot_wider(names_from = time_period, values_from = -c(area, time_period), names_sep = "_")

print(names(DF_wide))

head(DF_wide)

```

```{r}

set.seed(123) 

# function to simulate data based on mean and standard deviation
simulate_data <- function(mean_df, num_samples) {
  simulated_data <- data.frame()
  
  for (i in 1:nrow(mean_df)) {
    # extracting columns for time period t1 and t2 directly by names
    mean_t1 <- mean_df[i, which(grepl("_t1_mean$", names(mean_df)))]
    std_t1 <- mean_df[i, which(grepl("_t1_sd$", names(mean_df)))]
    mean_t2 <- mean_df[i, which(grepl("_t2_mean$", names(mean_df)))]
    std_t2 <- mean_df[i, which(grepl("_t2_sd$", names(mean_df)))]
    
    mean_t1 <- as.numeric(mean_t1)
    std_t1 <- as.numeric(std_t1)
    mean_t2 <- as.numeric(mean_t2)
    std_t2 <- as.numeric(std_t2)
    
    # extracting base names for the columns
    mean_names_t1 <- sub("_t1_mean$", "", names(mean_df)[which(grepl("_t1_mean$", names(mean_df)))])
    mean_names_t2 <- sub("_t2_mean$", "", names(mean_df)[which(grepl("_t2_mean$", names(mean_df)))])
    
    country <- mean_df$area[i]
    
    # simulate data for time period t1
    data_t1 <- data.frame(matrix(nrow = num_samples, ncol = length(mean_t1)))
    colnames(data_t1) <- paste0(mean_names_t1, "_t1")
    for (j in seq_along(mean_t1)) {
      data_t1[, j] <- rnorm(num_samples, mean = mean_t1[j], sd = std_t1[j])
    }
    
    # simulate data for time period t2
    data_t2 <- data.frame(matrix(nrow = num_samples, ncol = length(mean_t2)))
    colnames(data_t2) <- paste0(mean_names_t2, "_t2")
    for (j in seq_along(mean_t2)) {
      data_t2[, j] <- rnorm(num_samples, mean = mean_t2[j], sd = std_t2[j])
    }
    
    # combine the data for both periods
    combined_data <- cbind(data_t1, data_t2)
    combined_data$area <- country
    
    # add back floods and percipitation with the same values for each country
    combined_data$floods <- mean_df$floods_t1[i]
    combined_data$percipitation <- mean_df$percipitation_t1[i]
    
    simulated_data <- rbind(simulated_data, combined_data)
  }
  return(simulated_data)
}

# simulate data
simulated_data <- simulate_data(DF_wide, 15)  # 15 observations per country

head(simulated_data)

```


```{r}

data_scaled_ml <- as.data.frame(lapply(DF[, !(names(DF) %in% c("area"))], scale))

```

Free Model
```{r}

# free model with unique coefficients for each time period
model_free <- '
  # Time 1
  # Latent variables
  NutrientUseEfficiency_t1 =~ nue_t1 + pue_t1 + kue_t1
  AgriculturalEmissions_t1 =~ manure_direct_t1 + farm_energy_n20_t1 + enteric_fermentation_t1
  
  # Direct effects
  cereal_yield_t1 ~ c1*NutrientUseEfficiency_t1 + c2*pesticides_t1 + c3*gdp_capita_t1 + c4*AgriculturalEmissions_t1 + c5*percipitation_t1 + c6*temp_change_t1 + c7*employment_ratio_t1 + c8*water_withdrawal_t1
  
  # Mediator paths
  NutrientUseEfficiency_t1 ~ a1*pesticides_t1
  
  # Indirect effect calculations for mediated paths
  indirect1_t1 := a1*c1
  indirect2_t1 := a2*c1
  indirect3_t1 := a3*c1
  total_effect_t1 := c1 + indirect1_t1 + indirect2_t1 + indirect3_t1
  
  # Time 2
  # Latent variables
  NutrientUseEfficiency_t2 =~ nue_t2 + pue_t2 + kue_t2
  AgriculturalEmissions_t2 =~ manure_direct_t2 + farm_energy_n20_t2 + enteric_fermentation_t2
  
  # Direct effects
  cereal_yield_t2 ~ d1*NutrientUseEfficiency_t2 + d2*pesticides_t2 + d3*gdp_capita_t2 + d4*AgriculturalEmissions_t2 + d5*percipitation_t2 + d6*temp_change_t2 + d7*employment_ratio_t2 + d8*water_withdrawal_t2
  
  # Mediator paths
  NutrientUseEfficiency_t2 ~ e1*pesticides_t2 + e2*floods_t2 + e3*percipitation_t2
  
  # Indirect effect calculations for mediated paths
  indirect1_t2 := e1*d1
  indirect2_t2 := e2*d1
  indirect3_t2 := e3*d1
  total_effect_t2 := d1 + indirect1_t2 + indirect2_t2 + indirect3_t2
  
'

fit_free <- sem(model_free, data = data_scaled_ml)

```

Constrained model
```{r}

# Constrained model with common coefficients across both time periods
model_constrained <- '
  # Time 1
  # Latent variables
  NutrientUseEfficiency_t1 =~ nue_t1 + pue_t1 + kue_t1
  AgriculturalEmissions_t1 =~ manure_direct_t1 + farm_energy_n20_t1 + enteric_fermentation_t1
  
  # Direct effects
  cereal_yield_t1 ~ c*NutrientUseEfficiency_t1 + d*pesticides_t1 + e*gdp_capita_t1 + f*AgriculturalEmissions_t1 + g*percipitation_t1 + h*temp_change_t1 + i*employment_ratio_t1 + j*water_withdrawal_t1
  
  # Mediator paths
  NutrientUseEfficiency_t1 ~ k*pesticides_t1 + l*floods_t1 + m*percipitation_t1
  
  # Indirect effects through Nutrient Use Efficiency
  indirect_cereal_yield_t1 := (k*c) + (l*c) + (m*c)

  # Total effect
  total_effect_t1 := c + indirect_cereal_yield_t1
  
  # Time 2
  # Using the same structure as Time 1
  NutrientUseEfficiency_t2 =~ nue_t2 + pue_t2 + kue_t2
  AgriculturalEmissions_t2 =~ manure_direct_t2 + farm_energy_n20_t2 + enteric_fermentation_t2
  
  # Direct effects are constrained to be the same as Time 1
  cereal_yield_t2 ~ c*NutrientUseEfficiency_t2 + d*pesticides_t2 + e*gdp_capita_t2 + f*AgriculturalEmissions_t2 + g*percipitation_t2 + h*temp_change_t2 + i*employment_ratio_t2 + j*water_withdrawal_t2
  
  # Mediator paths are identical
  NutrientUseEfficiency_t2 ~ k*pesticides_t2 + l*floods_t2 + m*percipitation_t2
  
  # Indirect effects through Nutrient Use Efficiency
  indirect_cereal_yield_t2 := (k*c) + (l*c) + (m*c)

  # Total effect
  total_effect_t2 := c + indirect_cereal_yield_t2
'

fit_constrained <- sem(model_constrained, data = data_scaled_ml)

```

```{r}

multi_comparison <- lavTestLRT(fit_free, fit_constrained)
print(multi_comparison)

```