---
title: "Data Preparation"
author: "Leonor Brito"
date: "2024-09-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)

```

# Import data

```{r}
# function for importing and renaming columns
import_and_clean_data <- function(file, 
                                  rename_col = NULL, rename_to = NULL, 
                                  filter_item = NULL, drop_item = NULL, 
                                  filter_element = NULL, filter_year = NULL) {
  
  # import the dataset
  df <- read.csv(file)
  
  # rename column if specified
  if (!is.null(rename_col) && !is.null(rename_to)) {
    df <- df %>% rename(!!rename_to := !!sym(rename_col))
  }
  
  # filter rows based on Item values if specified
  if (!is.null(filter_item)) {
    df <- df %>% filter(Item %in% filter_item)
  }
  
  # drop specific Item rows if specified
  if (!is.null(drop_item)) {
    df <- df %>% filter(!Item %in% drop_item)
  }
  
  # filter rows based on Element values if specified
  if (!is.null(filter_element)) {
    df <- df %>% filter(Element %in% filter_element)
  }
  
  # filter rows based on Year if specified
  if (!is.null(filter_year)) {
    df <- df %>% filter(Year >= filter_year)
  }

  return(df)
}

```

Import all datasets using function above (note: a lot of these datasets were not used in the final model.) All datasets below were sourced from FAOSTAT (https://www.fao.org/faostat/en/#data) and AQUASTAT (https://data.apps.fao.org/aquastat). 
```{r}

# Yield (outcome variable)
yield <- import_and_clean_data("data/yield.csv")

# Socioeconomic data
credit <- import_and_clean_data("data/credit.csv")
employment <- import_and_clean_data("data/employment2.csv", rename_col = "Indicator", rename_to = "Item")
gdp_capita <- import_and_clean_data("data/gdp_capita.csv")
capital_stock <- import_and_clean_data("data/capital_stocks.csv")

# Environmental data
emissions <- import_and_clean_data("data/emissions.csv", drop_item = c("Drained organic soils (CO2)", "Drained organic soils (N2O)"))
temp_change <- import_and_clean_data("data/temp_change.csv", rename_col = "Element", rename_to = "Item")
aquastat <- readxl::read_excel('data/aquastat.xls')  # Imported as Excel file
LULUCF <- import_and_clean_data("data/LULUCF.csv", filter_item = c("Net Forest conversion"))

# Agricultural data
land_use <- import_and_clean_data("data/land_use.csv", drop_item = c("Agriculture area actually irrigated"))
pesticide_use <- import_and_clean_data("data/pesticide_use.csv")
irrigation <- import_and_clean_data("data/irrigation.csv")

# NUE data
nue <- import_and_clean_data("data/nue.csv", filter_element = c("Cropland nitrogen use efficiency", "Cropland phosphorus use efficiency", "Cropland potassium use efficiency"))
nue <- nue %>% select(-Item) %>% rename(Item = Element)

# Metadata
metadata <- import_and_clean_data("data/metadata.csv")

# Additional datasets 
pesticides <- import_and_clean_data("data/pesticides_added.csv")
expenditure <- import_and_clean_data("data/expenditure.csv")
emissions_new <- import_and_clean_data("data/emissions_added.csv")
emissions_added <- emissions_new %>%
  mutate(Element_Item = paste(Item, Element, sep = " - ")) %>%
  select(-c(Element, Item)) %>%
  rename(Item = Element_Item)
aquastat_added <- import_and_clean_data("data/aquastat_added.csv", rename_col = "Variable", rename_to = "Item") # sourced from AQUASTAT
labour <- import_and_clean_data("data/labour.csv", rename_col = "Indicator", rename_to = "Item")
investment <- import_and_clean_data("data/investment.csv")
livestock <- import_and_clean_data("data/livestock.csv")

# Livestock products
livestock_products <- read.csv('data/livestock_products.csv')
livestock_products <- aggregate(Value ~ Area + Year, data = livestock_products, sum)
livestock_products$Item <- "Yield"

# Additional data
yields <- import_and_clean_data("data/yield_details.csv")
gdp <- import_and_clean_data("data/gdp.csv")

```

# Data Prep

Function to change from long to wide format
```{r}

long_to_wide <- function(data) {
  wide_data <- pivot_wider(data, 
                           names_from = Item, 
                           values_from = Value, 
                           id_cols = c(Area, Year))
  return(wide_data)
}

```

Convert long to wide
```{r}

credit_wide <- long_to_wide(credit)
emissions_wide <- long_to_wide(emissions)
LULUCF_wide <- long_to_wide(LULUCF)
employment_wide <- long_to_wide(employment)
gdp_capita_wide <- long_to_wide(gdp_capita)
land_use_wide <- long_to_wide(land_use)
pesticide_use_wide <- long_to_wide(pesticide_use)
temp_change_wide <- long_to_wide(temp_change)
irrigation_wide <- long_to_wide(irrigation)
capital_stock_wide <- long_to_wide(capital_stock)
nue_wide <- long_to_wide(nue)
yield_wide <- long_to_wide(yield)
pesticides_wide <- long_to_wide(pesticides)
emissions_added_wide <- long_to_wide(emissions_added)
aquastat_added_wide <- long_to_wide(aquastat_added)
labour_wide <- long_to_wide(labour)
investment_wide <- long_to_wide(investment)
livestock_wide <- long_to_wide(livestock)
yields_wide <- long_to_wide(yields)
gdp_wide <- long_to_wide(gdp)
livestock_products_wide <- long_to_wide(livestock_products)

```

## Combine datasets

```{r}

combined_data <- yield_wide %>%
  left_join(yields_wide, by = c("Area", "Year")) %>%
  left_join(emissions_wide, by = c("Area", "Year")) %>%
  left_join(employment_wide, by = c("Area", "Year")) %>%
  left_join(gdp_capita_wide, by = c("Area", "Year")) %>%
  left_join(land_use_wide, by = c("Area", "Year")) %>%
  left_join(pesticide_use_wide, by = c("Area", "Year")) %>%
  left_join(temp_change_wide, by = c("Area", "Year")) %>%
  left_join(irrigation_wide, by = c("Area", "Year")) %>%
  left_join(capital_stock_wide, by = c("Area", "Year")) %>%
  left_join(nue_wide, by = c("Area", "Year")) %>%
  left_join(LULUCF_wide, by = c("Area", "Year")) %>%
  left_join(aquastat, by = "Area") %>%
  left_join(pesticides_wide, by = c("Area", "Year")) %>%
  left_join(emissions_added_wide, by = c("Area", "Year")) %>%
  left_join(aquastat_added_wide, by = c("Area", "Year")) %>%
  left_join(labour_wide, by = c("Area", "Year")) %>%
  left_join(investment_wide, by = c("Area", "Year")) %>%
  left_join(livestock_wide, by = c("Area", "Year")) %>%
  left_join(gdp_wide, by = c("Area", "Year")) %>%
  left_join(livestock_products_wide, by = c("Area", "Year"))

```

```{r}

# rename variables/columns
colnames(combined_data) <- c("area", "year", "cereal_yield", "barley_yield", "maize_yield", "wheat_yield","burning_residues", "drained_soils", 
  "employment_ag", "gdp_capita", "arable_land",
  "perm_meadows", "organic_ag", "pesticides", 
  "temp_change", "irrigation", "capital_stocks", 
  "nue", "pue", "kue", "forest_conv", "floods", 
  "percipitation", "insecticides", "herbicides",
  "fungicides", "rodenticides", "burning_residues_ch4", 
  "burning_residues_n20", "enteric_fermentation",
  "manure_ch4", "manure_direct", "manure_n20_direct", 
  "drained_soils_co2", "drained_soils_n20", 
  "farm_energy_ch4","farm_energy_n20", "farm_energy_co2",
  "rice_cultivation", "irrigation_equipped", "irrigated", 
  "grain_irrigation", "renewable_water_withdrawal", 
  "irrgation_efficiency", "water_efficiency", "water_withdrawal",
  "employment_rural", "labour", "employment_ratio", "employment_share",
  "FDI_inflows", "FDI_outflows", "livestock", "gdp", "livestock_products")

```

## Region Mapping
```{r}

# filter metadata to show only country code 5101 (EU countries)
filtered_metadata <- metadata %>%
  filter(Country.Group.Code == 5707)

# join the metadata with the combined_data
combined_data_region <- combined_data %>%
  right_join(filtered_metadata %>% select(Country, Country.Group.Code, Country.Group), by = c("area" = "Country"))

```

# Missing Data

```{r}

# check what variables are missing entirely for a given country
list_missing_by_country <- lapply(names(combined_data_region)[-1], function(var) {
  combined_data_region %>%
    group_by(area) %>%
    summarise(All_Missing = all(is.na(.data[[var]])), .groups = 'drop') %>%
    filter(All_Missing) %>%
    select(area) %>%
    mutate(Variable = var)
})

missing_by_country <- bind_rows(list_missing_by_country)

# print missing variables 
missing_by_country

```

## Visualize missingness

```{r}

# reshape dataset from wide to long format
df_long <- pivot_longer(
  combined_data_region,
  cols = -c(area, year, Country.Group, Country.Group.Code),
  names_to = "Variable",
  values_to = "Value"
)

# calculate the percentage of missing data per area (country) and year
percentage_missing <- df_long %>%
  group_by(area, year) %>%
  dplyr::summarize(Percent_Missing = mean(is.na(Value)) * 100) %>%
  ungroup()  

# create a heatmap using ggplot2
heatmap_plot <- ggplot(percentage_missing, aes(x = year, y = area, fill = Percent_Missing)) +
  geom_tile() + 
  scale_fill_gradient(low = "green", high = "red", name = "Percent Missing",
                      limits = c(0, 25)) +  
  theme_minimal() +
  labs(title = "Heatmap of Missing Data by Country and Year",
       x = "Year",
       y = "Country")

print(heatmap_plot)

```

```{r}

df_long <- gather(combined_data_region, key = "Variable", value = "Value", -area, -year, -Country.Group, -Country.Group.Code)

# calculate the percentage of missing data per year per variable
df_missing <- df_long %>%
  group_by(year, Variable) %>%
  dplyr::summarize(Percent_Missing = mean(is.na(Value)) * 100, .groups = 'drop')

# plot the time series of missing data percentage for each variable
ggplot(df_missing, aes(x = year, y = Percent_Missing, group = Variable)) +
  geom_line() +  
  facet_wrap(~ Variable, scales = "fixed") +  
  labs(title = "Time Series of Missing Data by Variable",
       x = "Year",
       y = "Percentage Missing") +
  theme_minimal()  

```

## Remove missing

Filter from 2005-2020
Remove Malta, Luxembourg and Ireland
Remove rodenticides and grain irrigation variables
```{r}

clean_df <- subset(combined_data_region,
                   year >= 2006 & year <= 2020 & 
                   !area %in% c("Malta", "Belgium-Luxembourg"),
                   select = -c(grain_irrigation, rice_cultivation, FDI_inflows, FDI_outflows, employment_share, irrigation, rodenticides, maize_yield))

```



```{r}

save(clean_df, file = "clean_df.RData")

```

