---
title: "Exploratory Data Analysis"
author: "Leonor Brito"
date: "2024-09-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages
```{r}

library(lavaan)
library(corrplot)
library(Hmisc)
library(ggplot2)
library(purrr)
library(dplyr)

```

# Exploratory Data Analysis

```{r}

load("/Users/leonorbrito/Documents/Master/Thesis/Code/clean_df.RData")

```


## Correlation

First, a correlation matrix was created using all potentially relevant variables. 
```{r}

all_vars <- clean_df[, 3:(ncol(clean_df) - 2)]

corr_results <- rcorr(as.matrix(all_vars))

# access the matrix of correlation coefficients
cor_matrix <- corr_results$r

# access the matrix of p-values
p_matrix <- corr_results$P

# visualize the correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, tl.cex = 0.6, number.cex = 0.5, addCoef.col = "black")

# create a correlation plot with significance masking
corrplot(cor_matrix, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45, sig.level = 0.05, insig = "blank", addCoef.col = "black", number.cex = 0.5, tl.cex = 0.6)

```

## Normality / Outlier Detection

```{r}

# function to create histogram for each variable to check for normality/skewness
plot_distribution <- function(data, variable_name) {
  p1 <- ggplot(data, aes_string(x = variable_name)) + 
        geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
        ggtitle(paste("Histogram of", variable_name))

  # print histogram
  print(p1)
}

apply_checks <- function(data) {
  numeric_columns <- sapply(data, is.numeric)  # Identify numeric columns
  for (variable_name in names(numeric_columns[numeric_columns])) {
    plot_distribution(data, variable_name)     # Plot histograms for each variable
  }
}

# apply normality check for all variables
apply_checks(all_vars)

```


Together with the findings of the correlation matrix and the normality checks, a few different SEM model structures were explored to experiment with various combinations of variables, relationships and their impact on model fit. This was purely experimental and provided insights into what variables to consider / which ones to remove from the study to avoid multi-collinearity and skewed data. 

## Experimental model structures

Scaling

```{r}

data_sem <- clean_df[, !(names(clean_df) %in% c("Country.Group.Code", "Country.Group", "gdp_category"))]
country_data <- data_sem[,c("area"), drop = FALSE]
data_to_scale <- data_sem[, !(names(data_sem) %in% c("area"))]
data_scaled <- as.data.frame(lapply(data_to_scale, scale))

data_for_sem <- cbind(data_scaled, country_data)

```


```{r}

# model 1: includes one latent variable (NUE)

model_1 <- '
  NutrientUseEfficiency =~ 1*nue + pue + kue
  cereal_yield ~ NutrientUseEfficiency + pesticides + percipitation + gdp_capita + temp_change + irrgation_efficiency
  NutrientUseEfficiency ~ pesticides + burning_residues
'

# fit the model
fit_model_1 <- sem(model_1, data = data_for_sem, cluster = 'area', missing = "fiml")
summary(fit_model_1, fit.measures=TRUE)

```

```{r}

## model 2: adding more pesticide types (herbicides, fungicides, etc.) influencing cereal yield and NUE
model_2 <- '
  NutrientUseEfficiency =~ 1*nue + pue + kue
  cereal_yield ~ NutrientUseEfficiency + gdp_capita + irrgation_efficiency + percipitation + herbicides + fungicides + rodenticides + manure_direct
  NutrientUseEfficiency ~ insecticides + fungicides + burning_residues_n20 + manure_direct + enteric_fermentation
'

# fit the model
fit_model_2 <- sem(model_2, data = data_for_sem, cluster = 'area', missing = "fiml")
summary(fit_model_2, fit.measures=TRUE)

```

```{r}

# model 3: simplifies pesticide types influencing cereal yield, focusing on herbicides
model_3 <- '
  NutrientUseEfficiency =~ 1*nue + pue + kue
  cereal_yield ~ NutrientUseEfficiency + gdp_capita + irrgation_efficiency + percipitation + herbicides
  NutrientUseEfficiency ~ burning_residues_n20 + manure_direct
'

# fit the model
fit_model_3 <- sem(model_3, data = data_for_sem, cluster = 'area', missing = "fiml")
summary(fit_model_3, fit.measures=TRUE)

```

```{r}

# model 4: adding two new latent variables, WaterUse and AgriculturalEmissions with a direct effect on yield
model_4 <- '
  NutrientUseEfficiency =~ 1*nue + pue + kue
  WaterUse =~ 1*water_withdrawal + irrgation_efficiency + renewable_water_withdrawal
  AgriculturalEmissions =~ 1*manure_direct + farm_energy_n20 + enteric_fermentation
  cereal_yield ~ NutrientUseEfficiency + pesticides + gdp_capita + WaterUse + AgriculturalEmissions + percipitation + temp_change
  NutrientUseEfficiency ~ pesticides + floods + percipitation
'

# fit the model
fit_model_4 <- sem(model_4, data = data_for_sem, cluster = 'area', missing = "fiml")
summary(fit_model_4, fit.measures=TRUE)

```

```{r}

# model 5: adding Economy as a new latent variable affecting cereal yield along with WaterUse and Agricultural Emissions
model_5 <- '
  NutrientUseEfficiency =~ 1*nue + pue + kue
  WaterUse =~ 1*water_withdrawal + irrgation_efficiency + renewable_water_withdrawal
  AgriculturalEmissions =~ 1*manure_direct + farm_energy_n20 + enteric_fermentation
  Economy =~ 1*gdp_capita + labour + employment_ratio
  cereal_yield ~ NutrientUseEfficiency + pesticides + Economy + WaterUse + AgriculturalEmissions + percipitation + temp_change
  NutrientUseEfficiency ~ pesticides + floods + percipitation
'

# fit the model
fit_model_5 <- sem(model_5, data = data_for_sem, cluster = 'area', missing = "fiml")
summary(fit_model_5, fit.measures=TRUE)


```

```{r}

# model 6: focusing on Agricultural Emissions and Water Use as latent variables influencing cereal yield and NUE
model_6 <- '
  NutrientUseEfficiency =~ nue + pue + kue
  AgriculturalEmissions =~ 1*burning_residues + manure_direct + farm_energy_n20 + enteric_fermentation
  cereal_yield ~ NutrientUseEfficiency + pesticides + gdp_capita + irrgation_efficiency + AgriculturalEmissions + percipitation + temp_change
  NutrientUseEfficiency ~ pesticides + floods + percipitation
'

# fit the model
fit_model_6 <- sem(model_6, data = data_for_sem, cluster = 'area', missing = "fiml")
summary(fit_model_6, fit.measures=TRUE)


```

```{r}

# model 8: adding year and employment ratio to the structure, testing temporal effects on yield and NUE
model_7 <- '
  NutrientUseEfficiency =~ 1*nue + pue + log_kue
  AgriculturalEmissions =~ 1*manure_direct + log_farm_energy_n20 + log_enteric_fermentation
  cereal_yield ~ NutrientUseEfficiency + pesticides + gdp_capita + water_withdrawal + percipitation + AgriculturalEmissions + employment_ratio + temp_change + year
  NutrientUseEfficiency ~ pesticides + floods + percipitation + year
'

# fit the model
fit_model_7 <- sem(model_7, data = data_for_sem, cluster = 'area', missing = "fiml", estimator = "MLR")
summary(fit_model_7, fit.measures=TRUE)


```


